---
title: "EPI 560: Lab 5"
author: "Alie Vandi-tarawally"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: paged
        theme: cosmo
        toc: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "build") })
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Setup
library(pacman)
p_load("tidyverse", "knitr", "DBI", "RSQLite", "sqldf")
```

## Lab 5: Responses 
### Part I
```{r part_1, echo=FALSE, message=FALSE, warning=FALSE}
datab <- dbConnect(SQLite(), dbname = "~/Documents/EPI560/epi560.sqlite")
dbListTables(datab)

#Q1. To count the number of participants

count_demo <- dbGetQuery(datab, "SELECT COUNT(*) 
AS PatientCount
FROM  P_DEMO")
kable(count_demo)

#Q2. Join demographics and sleep quality tables and count the number of observations in each table.

Demo_SD <- dbGetQuery(datab, "SELECT * FROM P_DEMO 
INNER JOIN P_SLQ
ON P_DEMO.SEQN = P_SLQ.SEQN")
kable(head(Demo_SD, 5))

count_Demo_SD <- dbGetQuery(datab, "SELECT COUNT(*) 
AS PatientCount
FROM P_DEMO 
INNER JOIN P_SLQ 
ON P_DEMO.SEQN = P_SLQ.SEQN;")
kable(count_Demo_SD)
```
The number of patients included in the demographic dataset is `r Demo_SD$PatientCount`.
The number of patients included in the merged dataset is `r count_Demo_SD$PatientCount`.
### Part II
```{r part_2, echo=FALSE, message=FALSE, warning=FALSE}
datab <- dbConnect(SQLite(), dbname = "~/Documents/EPI560/epi560.sqlite")
# Q.3 Some of the variables I would need for my final project
DEM_AUD_SD <- dbGetQuery(datab, "SELECT 
    P_DEMO.RIAGENDR,
    P_DEMO.RIDAGEYR,
    P_DEMO.DMDEDUC2,
    P_DEMO.DMDMARTZ,
    P_AUQ.AUQ054,
    P_AUQ.AUQ060,
    P_AUQ.AUQ070,
    P_AUQ.AUQ080,
    P_AUQ.AUQ090,
    P_SLQ.SLQ300,
    P_SLQ.SLQ310
FROM 
    P_DEMO
LEFT JOIN 
    P_AUQ ON P_DEMO.SEQN = P_AUQ.SEQN
LEFT JOIN 
    P_SLQ ON P_DEMO.SEQN = P_SLQ.SEQN
LIMIT 5")
kable(DEM_AUD_SD)

# Variables needed to respond to the questions in Lab5 Part II
DEM_AUD_SD_1 <- dbGetQuery(datab, "SELECT 
    P_DEMO.RIAGENDR,
    P_AUQ.AUQ054,
    P_AUQ.AUQ060,
    P_AUQ.AUQ070,
    P_AUQ.AUQ080,
    P_SLQ.SLQ300,
    P_SLQ.SLQ310
FROM 
    P_DEMO
LEFT JOIN 
    P_AUQ ON P_DEMO.SEQN = P_AUQ.SEQN
LEFT JOIN 
    P_SLQ ON P_DEMO.SEQN = P_SLQ.SEQN
LIMIT 5")
kable(DEM_AUD_SD_1)

# Q.4 Average usual sleep time among participants with good or excellent hearing
Gd_Ext <- dbGetQuery(datab, "SELECT 
    AVG(P_SLQ.SLQ310) AS Avg_SP_Time
FROM 
    P_AUQ
LEFT JOIN 
    P_SLQ ON P_AUQ.SEQN = P_SLQ.SEQN
WHERE 
    P_AUQ.AUQ054 IN (1, 2)  
    AND P_SLQ.SLQ310 IS NOT NULL")
kable(Gd_Ext)

# Average usual sleep time among participants with  a little trouble, moderate trouble, or a lot of trouble with their hearing
Mod_T <- dbGetQuery(datab, "SELECT 
    AVG(P_SLQ.SLQ310) AS Avg_SP_Time
FROM 
    P_AUQ
LEFT JOIN 
    P_SLQ ON P_AUQ.SEQN = P_SLQ.SEQN
WHERE 
    P_AUQ.AUQ054 IN (3, 4, 5)  
    AND P_SLQ.SLQ310 IS NOT NULL")
kable(Mod_T)

# Q.5. Repeat Q4 but with age variable added.
DEM_AUD_SD_5 <- dbGetQuery(datab, "SELECT 
    P_DEMO.RIDAGEYR,
    P_AUQ.AUQ054,
    P_AUQ.AUQ060,
    P_AUQ.AUQ070,
    P_AUQ.AUQ080,
    P_SLQ.SLQ300,
    P_SLQ.SLQ310
FROM 
    P_DEMO
LEFT JOIN 
    P_AUQ ON P_DEMO.SEQN = P_AUQ.SEQN
LEFT JOIN 
    P_SLQ ON P_DEMO.SEQN = P_SLQ.SEQN
LIMIT 6")
kable(DEM_AUD_SD_5)

# Estimate the average sleep time for participants with either good or excellent hearing.

Gd_Ext_5 <- dbGetQuery(datab, "SELECT 
    P_DEMO.RIDAGEYR AS Age,
    AVG(P_SLQ.SLQ310) AS Avg_SP_Time
FROM 
    P_AUQ
LEFT JOIN 
    P_SLQ ON P_AUQ.SEQN = P_SLQ.SEQN
LEFT JOIN 
    P_DEMO ON P_AUQ.SEQN = P_DEMO.SEQN
WHERE 
    P_AUQ.AUQ054 IN (1, 2)  
    AND P_SLQ.SLQ310 IS NOT NULL
    AND P_DEMO.RIDAGEYR IS NOT NULL
GROUP BY 
    P_DEMO.RIDAGEYR
ORDER BY 
    P_DEMO.RIDAGEYR
LIMIT 6")
kable(Gd_Ext_5)

# Estimate the average sleep time for participants with moderate to troubled hearing. 
Mod_T_5 <- dbGetQuery(datab, "SELECT 
    P_DEMO.RIDAGEYR AS Age,
    AVG(P_SLQ.SLQ310) AS Avg_SP_Time
FROM 
    P_AUQ
LEFT JOIN 
    P_SLQ ON P_AUQ.SEQN = P_SLQ.SEQN
LEFT JOIN 
    P_DEMO ON P_AUQ.SEQN = P_DEMO.SEQN
WHERE 
    P_AUQ.AUQ054 IN (3, 4, 5)  
    AND P_SLQ.SLQ310 IS NOT NULL
    AND P_DEMO.RIDAGEYR IS NOT NULL
GROUP BY 
    P_DEMO.RIDAGEYR
ORDER BY 
    P_DEMO.RIDAGEYR
LIMIT 6")
kable(Mod_T_5)

# Q.6. 
# Count missing values in the hearing disorder variables 
Miss_P_SLQ <- dbGetQuery(datab, "SELECT
  SUM(CASE WHEN AUQ054 IS NULL THEN 1 ELSE 0 END) AS missing_whisper,
  SUM(CASE WHEN AUQ060 IS NULL THEN 1 ELSE 0 END) AS missing_normal,
  SUM(CASE WHEN AUQ070 IS NULL THEN 1 ELSE 0 END) AS missing_shout
FROM P_AUQ
")
kable(Miss_P_SLQ)
# Stratify by age 50
Miss_P_SLQ_50 <- dbGetQuery(datab, "SELECT
  CASE WHEN d.RIDAGEYR < 50 THEN 'Under 50' ELSE '50 and over' END AS age,
  COUNT(*) AS total,
  SUM(CASE WHEN a.AUQ054 IS NULL THEN 1 ELSE 0 END) AS missing_whisper,
  SUM(CASE WHEN a.AUQ060 IS NULL THEN 1 ELSE 0 END) AS missing_normal,
  SUM(CASE WHEN a.AUQ070 IS NULL THEN 1 ELSE 0 END) AS missing_shout
FROM P_AUQ a
JOIN P_Demo d ON a.SEQN = d.SEQN
GROUP BY age")
kable(Miss_P_SLQ_50)

# Stratify by sleep time over 8 hours
Miss_P_SLQ_SP <- dbGetQuery(datab, "SELECT
  CASE WHEN s.SLQ310 > 8 THEN 'Over 8 hrs' ELSE '8 or fewer hrs' END AS sleep,
  COUNT(*) AS total,
  SUM(CASE WHEN a.AUQ054 IS NULL THEN 1 ELSE 0 END) AS missing_whisper,
  SUM(CASE WHEN a.AUQ060 IS NULL THEN 1 ELSE 0 END) AS missing_normal,
  SUM(CASE WHEN a.AUQ070 IS NULL THEN 1 ELSE 0 END) AS missing_shout
FROM P_AUQ a
JOIN P_SLQ s ON a.SEQN = s.SEQN
GROUP BY sleep")
kable(Miss_P_SLQ_SP)

# Note: I used chatgpt for this question. I copied the question and pasted in chatgpt with the prompt "I am struggling to understand the instruction in this question. Please unpack for me and explain how to execute it in DB Browser."

# Q.7. Create a dataframe of participants with either good or excellent hearing who usually sleeps more than 8 hours per week

SLQ_Good_8 <- dbGetQuery(datab, "SELECT P_AUQ.SEQN, P_AUQ.AUQ054, P_SLQ.SLQ310
FROM P_AUQ AS P_AUQ
JOIN P_SLQ AS P_SLQ ON P_AUQ.SEQN = P_SLQ.SEQN
WHERE P_AUQ.AUQ054 IN (1, 2)
  AND P_SLQ.SLQ310 > 8
LIMIT 6")
kable(SLQ_Good_8)
```
### Q4
The average sleep time among participants with either good or excellent hearing is `r Gd_Ext$Avg_SP_Time`.

The average sleep time among participants with  a little trouble, moderate
trouble, or a lot of trouble with their hearing is `r Mod_T$Avg_SP_Time`.

### Q5. 
The average sleep time among the first six participants with either good or excellent hearing grouped by age are `r Gd_Ext_5$Avg_SP_Time`.

The average sleep time among the first six participants with  a little trouble, moderate
trouble, or a lot of trouble with their hearing grouped by age are `r Mod_T_5$Avg_SP_Time`.

### Q6.
Since the percentage of missingness greatly differ between the groups, the missingness is differential.

### Q7
The six participants with either good or excellent hearing who usually sleeps more than 8 hours per week are `r SLQ_Good_8$SLQ310`.



\pagebreak  

## Code Appendix  
 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE} 
```
