---
title: "EPI 560: Lab 8"
author: "Alie Vandi-tarawally"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: paged
        theme: cosmo
        toc: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "build") })
---
## Setup
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(pacman)
p_load("tidyverse", "knitr", "lubridate")
```

## Load the dataset 
```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
getwd()
df <- read_csv("/Users/alievandi-tarawally/Downloads/lab8_data.csv")
dplyr::tbl_df(df)
utils::View(df)
```
## Lab 8: Responses  
### Part 1
```{r part_1, echo=FALSE, message=FALSE, warning=FALSE}
# Create a new column called dx_date_local

df <- df %>%
  mutate(dx_date = as.POSIXct(dx_date, 
                              format = "%Y-%m-%d %H:%M:%S", 
                              tz = "UTC"))

df$dx_date_local = as.POSIXct(df$dx_date)

for (i in 1:nrow(df)) {
  df$dx_date_local[i] <- force_tz(df$dx_date[i], 
                                  tzone = df$time_zone[i])
}
dplyr::tbl_df(df)
```
## Convert local timezones into UTC
```{r}
df <- df %>%
  rowwise() %>%
  mutate(dx_date_utc = with_tz(dx_date_local, tzone = "UTC")) %>%
  ungroup()

head(df, 10)

```
After running the code above, I noticed there was no difference between the local and UTC timezones. To investigate further, I checked the local timezone and found that it was actually set to UTC. As a result, I correctly assigned the local timezones and then converted the times to UTC.

```{r}
# Check for the correct timezone of dx_date_local
attr(df$dx_date_local, "tzone")

# Correct local time zones
df <- df %>%
  rowwise() %>%
  mutate(dx_date_local = force_tz(dx_date_local, tzone = time_zone)) %>%
  ungroup()

# Convert from local to UTC
df <- df %>%
  rowwise() %>%
  mutate(dx_date_utc = with_tz(dx_date_local, tzone = "UTC")) %>%
  ungroup()

head(df, 5)

```
### Part 2
#### code for part 2
```{r part_2, echo=FALSE, message=FALSE, warning=FALSE}
df_long <- df %>%
  group_by(id) %>%
  arrange(dx_date) %>%
  mutate(rank = row_number()) %>%
  ungroup()

kable(head(df_long, 20))
```
b. The rank column assigns a sequential number to each event for the same patient. For example, when a patient experiences an infection for the first time, it is ranked as 1; the second infection is ranked as 2, and so on. In the table above, the patient with ID 138 first had COVID-19 on 2019-01-06 at 11:44:00 (UTC), which was assigned a rank of 1. The same patient later had Mumps on 2019-01-07 at 04:47:00 (UTC), which was assigned a rank of 2.

```{r}
df_wide <- df_long %>%
  select(id, rank, dx, dx_date) %>%
  pivot_wider(
    names_from = rank,
    values_from = c(dx, dx_date),
    names_glue = "{.value}_{rank}"
  )

print(head(df_wide, 5))

# Check if number of rows match number of unique IDs
nrow(df_wide)
n_distinct(df$id)
```
### Part 3
#### code for part 3

```{r}
df_wide <- df_wide %>%
  mutate(
    
    time_between_dxs = as.numeric(difftime(dx_date_2, dx_date_1, units = "days")),
   
    time_cat = case_when(
      time_between_dxs <= 1 ~ "0-1 days",
      time_between_dxs <= 30 ~ "2-30 days",
      time_between_dxs <= 180 ~ "31-180 days",
      time_between_dxs <= 365 ~ "181-365 days",
      TRUE ~ ">365 days"
    )
  )

head(df_wide)

nrow(df_wide)
```
```{r}
# Create a new column for patients with only one diagnosis
df_wide <- df_wide %>%
  mutate(
    time_cat_with_one = ifelse(is.na(dx_2), "One diagnosis", time_cat)
  )

# Summarize the number of patients in each category, including the category for only one diagnosis.
summary_df <- df_wide %>%
  group_by(time_cat_with_one) %>%
  summarise(patient_count = n_distinct(id)) %>%
  arrange(time_cat_with_one)

kable(summary_df, format = 'pipe', col.names = c("Category", "Number of Patients"))
```







\pagebreak  

## Code Appendix  
 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE} 
```
