---
title: "EPI 560: Lab9"
author: "Alie Malcolm Vandi-tarawally"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: paged
        theme: cosmo
        toc: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "build") })
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Setup
library(pacman)
p_load("tidyverse", "knitr", "DBI", "RSQLite", "summarytools")
```

## Lab {X}: Responses  
```{r part_1, echo=FALSE, message=FALSE, warning=FALSE}
# code for part 1
rm(list = ls())
data <- data.frame(
id = c(12325, 12325, 29080, 33422, 33422, 33422, 33422, 43245, 50989, 50989),
county_fips = c(06001, 06001, 06041, 06055, 06055, 06055, 06055, 06097, 06075, 06075),
state_fips = c(06,06,06,06,06,06,06,06,06,06),
walking_minutes = rnorm(10, 30, 10)
)
head(data)

# for loop
unique_ids <- unique(data$id)
data$cumulative_walking <- NA
for (i in unique_ids) {
cumulative_walking_for_id <- sum(data$walking_minutes[data$id == i]) %>% round(1)
data$cumulative_walking[data$id == i] <- round(cumulative_walking_for_id, 1)
message <- paste0("id", i, ": cumulative walking time = ",
cumulative_walking_for_id, " minutes.")
print(message)
}

# while loop
i <- 1
count <- 0
threshold <- 60
while (i <= nrow(data)) {
if (data$cumulative_walking[i] > threshold) {count <- count + 1
print(paste0("entry exceeds threshold, current count = ", count))
} else {
}
i <- i + 1
}

# write another while loop example
i <- 1
count <- 0
threshold <- 60

# Print the count of entries exceeding the threshold
print(paste0("number of entries exceeding the threshold: ", count))
```
```{r}
pad_fips <- function(data, cols) {
data <- data %>%
mutate(across(all_of(cols),
~ifelse(nchar(.) == 4 | nchar(.) == 1,
paste0("0", .), .)))
return(data)
}
columns_to_pad <- c("county_fips", "state_fips")
new_data <- pad_fips(data, columns_to_pad)
head(new_data)
```
### 1. Load NHANES database and create a dataframe
```{r}
datab <- dbConnect(SQLite(), dbname = "~/Documents/EPI560/epi560.sqlite")

query <- "
SELECT *
FROM P_SLQ
LEFT JOIN P_AUQ
ON P_SLQ.SEQN = P_AUQ.SEQN
"

SLQ_AUQ_Merged <- dbGetQuery(datab, query)

SLQ_AUQ_Merged
```
2. For this analysis, all responses that are marked as "refused to answer," "don't know," missing, or invalid will be treated as missing values (NAs). This approach allows for the consistent handling of non-informative or unreliable responses. By treating these entries as NAs, we avoid discarding entire observations, which helps preserve valuable data while maintaining the integrity of the analysis. We believe this will make our analysis accurate by reducing bias from excluding them or not accounting for them. 

### 3.
```{r}
cleaned_data <- SLQ_AUQ_Merged 

missing_values <- c("77777", "99999", "")

cleaned_data$SLQ300[cleaned_data$SLQ300 %in% missing_values] <- NA

cleaned_data$SLQ310[cleaned_data$SLQ310 %in% missing_values] <- NA

# Print frequency tables for SLQ300 and SLQ310 after recoding
after_cleaning <- cleaned_data[, c("SLQ300", "SLQ310")]

print(freq(after_cleaning))
```
### 4.
```{r}
cleaned_sleep_data <- SLQ_AUQ_Merged

vars_to_clean <- c("SLQ300", "SLQ310", "SLQ320", "SLQ330")

missing_codes <- c("77777", "99999", "")

# Recode loop
for (var in vars_to_clean) {
  cleaned_sleep_data[[var]][cleaned_sleep_data[[var]] %in% missing_codes] <- NA
}

freq(cleaned_sleep_data$SLQ320)
```
The for loop improved the process by avoiding manually writing the codes repetitively. 
NHANES should maintain standardization of "missing" and "don't know" values to ensure consistent and accurate data cleaning, analysis and interpretation.
5. 
**Variables for analysis**
- SLQ dataset
  -- SLD012 (Sleep hours - weekdays or workdays)
  -- SLD013 (Sleep hours - weekends)
- AUQ dataset
  -- AUQ054 (General condition of hearing)
  -- AUQ060 (Hear a whisper from across a quiet room?)
  -- AUQ070 (Hear normal voice across a quiet room?)
  -- AUQ410d (Cause of hearing loss-Illness/Infections)
- DEMO
  -- RIAGENDR (Gender)
  -- RIDAGEYR (Age in years at screening)


### 6. 
```{r}
query <- "
SELECT 
  P_PAQ.PAQ610,
  P_ALQ.ALQ121,
  P_GHB.LBXGH,
  P_AUQ.AUQ054,
  P_SLQ.SLQ050
FROM P_PAQ
LEFT JOIN P_ALQ ON P_PAQ.SEQN = P_ALQ.SEQN
LEFT JOIN P_GHB ON P_PAQ.SEQN = P_GHB.SEQN
LEFT JOIN P_AUQ ON P_PAQ.SEQN = P_AUQ.SEQN
LEFT JOIN P_SLQ ON P_PAQ.SEQN = P_SLQ.SEQN
"
selected_nhanes_data <- dbGetQuery(datab, query)

head(selected_nhanes_data)

```

```{r}
cleaned_nhanes <- selected_nhanes_data

# Convert variables to character to ensure consistent handling of special codes
cleaned_nhanes$PAQ610 <- as.character(cleaned_nhanes$PAQ610)
cleaned_nhanes$ALQ121 <- as.character(cleaned_nhanes$ALQ121)
cleaned_nhanes$LBXGH  <- as.character(cleaned_nhanes$LBXGH)
cleaned_nhanes$AUQ054 <- as.character(cleaned_nhanes$AUQ054)
cleaned_nhanes$SLQ050 <- as.character(cleaned_nhanes$SLQ050)

# Clean variables: recode 77, 99, and "." as NA
cleaned_nhanes$PAQ610[cleaned_nhanes$PAQ610 %in% c("77", "99", ".")] <- NA

cleaned_nhanes$ALQ121[cleaned_nhanes$ALQ121 %in% c("77", "99", ".")] <- NA

cleaned_nhanes$LBXGH[cleaned_nhanes$LBXGH %in% c(".")] <- NA

cleaned_nhanes$AUQ054[cleaned_nhanes$AUQ054 %in% c("77", "99", ".")] <- NA

cleaned_nhanes$SLQ050[cleaned_nhanes$SLQ050 %in% c("7", "9", ".")] <- NA

# Show the first six rows of the cleaned dataset
kable(head(cleaned_nhanes[, c("PAQ610", "ALQ121", "LBXGH", "AUQ054", "SLQ050")]))

```
### 7. 
```{r}
# Function to cap continuous variables at 5th and 95th percentiles
clean_continuous_vars <- function(data, vars) {
  for (var in vars) {
    # Convert to numeric
    data[[var]] <- as.numeric(data[[var]])
    
    # Calculate 5th and 95th percentiles, ignoring NA
    p5 <- quantile(data[[var]], 0.05, na.rm = TRUE)
    p95 <- quantile(data[[var]], 0.95, na.rm = TRUE)
    
    # Cap values outside of range
    data[[var]] <- ifelse(data[[var]] < p5, p5,
                          ifelse(data[[var]] > p95, p95, data[[var]]))
  }
  return(data)
}

# Function to recode categorical variables: 99 → 9999
clean_categorical_vars <- function(data, vars) {
  data <- data %>%
    mutate(across(all_of(vars), ~ ifelse(. == "99", "9999", .)))
  return(data)
}

nhanes_data <- selected_nhanes_data %>%
  mutate(across(c("ALQ121", "SLQ050", "AUQ054"), as.character),
         across(c("PAQ610", "LBXGH"), as.numeric))

continuous_vars <- c("PAQ610", "LBXGH")
categorical_vars <- c("ALQ121", "SLQ050", "AUQ054")

# Apply cleaning functions
cleaned_nhanes <- nhanes_data %>%
  clean_continuous_vars(continuous_vars) %>%
  clean_categorical_vars(categorical_vars)

# View cleaned result
kable(head(cleaned_nhanes[, c(continuous_vars, categorical_vars)]))
```

### 2nd part of Q7. 
```{r}
# Function to cap continuous variables at 5th and 95th percentiles
clean_continuous_vars <- function(data, vars) {
  for (var in vars) {
    # Convert to numeric
    data[[var]] <- as.numeric(data[[var]])
    
    # Calculate 5th and 95th percentiles, ignoring NA
    p5 <- quantile(data[[var]], 0.05, na.rm = TRUE)
    p95 <- quantile(data[[var]], 0.95, na.rm = TRUE)
    
    # Cap values outside the range
    data[[var]] <- ifelse(data[[var]] < p5, p5,
                          ifelse(data[[var]] > p95, p95, data[[var]]))
  }
  return(data)
}

# Function to recode categorical variables: 99 → 9999
clean_categorical_vars <- function(data, vars) {
  data <- data %>%
    mutate(across(all_of(vars), ~ ifelse(. == "99", "9999", .)))
  return(data)
}

original_nhanes <- selected_nhanes_data

nhanes_data <- original_nhanes %>%
  mutate(across(c("ALQ121", "SLQ050", "AUQ054"), as.character),
         across(c("PAQ610", "LBXGH"), as.numeric))

continuous_vars <- c("PAQ610", "LBXGH")
categorical_vars <- c("ALQ121", "SLQ050", "AUQ054")

cleaned_nhanes <- nhanes_data %>%
  clean_continuous_vars(continuous_vars) %>%
  clean_categorical_vars(categorical_vars)

kable(head(cleaned_nhanes[, c(continuous_vars, categorical_vars)]))


# Histogram of original LBXGH
ggplot(original_nhanes, aes(x = as.numeric(LBXGH))) +
  geom_histogram(binwidth = 0.2, fill = "skyblue", color = "black") +
  labs(title = "Original Glycohemoglobin % (LBXGH)", x = "LBXGH", y = "Frequency")

# Histogram of cleaned LBXGH
ggplot(cleaned_nhanes, aes(x = as.numeric(LBXGH))) +
  geom_histogram(binwidth = 0.2, fill = "darkgreen", color = "black") +
  labs(title = "Cleaned Glycohemoglobin % (LBXGH)", x = "LBXGH", y = "Frequency")

```



\pagebreak  

## Code Appendix  
 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE} 
```
